Runtime-Assisted Convergence in Replicated Data Types

### 2022-03-21 introduction.tex
Edits for shepherd review and camera-ready

Large-scale web services and decentralized applications often rely on
geo-distributed state replication to meet their latency and availability
needs. An application's state is replicated asynchronously in such a
setting, meaning that operations are independently executed at each
replica, and updates are applied asynchronously at other replicas after a
possible network delay. Asynchronous execution complicates programming and
reasoning about distributed applications as it induces the possibility of
conflicting updates leading to non-convergence and application integrity
violations. Two basic approaches have been proposed to address this
problem. The first approach is to selectively strengthen the system
consistency to pre-empt the conflicting updates. The second is to redefine
the application state in terms of \emph{Replicated Data Types} (RDTs) that
are specially engineered to handle conflicting updates. Strengthening
system consistency necessarily entails inter-replica coordination,
therefore applications prefer to utilize RDTs as much as possible,
resorting to consistency strengthening only when it is necessary to maintain
application integrity. The common design principle guiding the development
of replicated data types is commutativity. The idea is that if the
replicated state is only updated by commutative operations, then updates
can be applied in any order and the replica states are still guaranteed to
converge. Indeed, there exist common usecases in web applications that can
be implemented using Commutative Replicated Data Types
(CRDTs)~\cite{crdts}. For instance, a video view counter on a streaming
application such as YouTube can be implemented as a Counter CRDT that
supports commutative increments.

In general, however, commutativity is not a common occurrence among
data types. Most common data type definitions come with at least a
pair of operations that do not commute. For instance an \C{add} and a
\C{remove} operations on a set do not commute if both are for the same
element.  Likewise two \C{insert} operations for the same position in
a list do not commute. Such non-commutative operations translate to
conflicting updates in a distributed setting, whose cumulative result
cannot be determined by the sequential specification of the data type
alone. To use a non-commutative data type in a replicated setting,
creative re-engineering of its internal representation and algorithms
is needed to turn it into a bonafide CRDT with a sensible distributed
semantics.
% In the case of set data type, that could involve changing the internal
% representation to maintain two sets -- a set containing the current
% elements and a \emph{tombstone} set that contains every element that
% has ever been deleted from the set. A \C{remove $k$} operation removes
% $k$ from the element set and adds it to the tombstone set, and an
% \C{insert $k$} operation adds $k$ to the element set only if it is not
% present in the tombstone set (i.e., only if $k$ was not previously
% deleted from the set). This results in a CRDT set called
% \emph{two-phase set} where an element once removed cannot be re-added. 
% A more complex re-engineering of the set data type could result in a set
% CRDT with a different (and arguably more useful) semantics.
% Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
% allows elements to be readded, but which prioritizes \C{remove}s over
% concurrent \C{insert}s. This transformation requires vector
% clocks~\cite{vectorclock} to keep track of casual ordering among
% \C{insert}s and \C{remove}s.
Indeed, most CRDTs have such carefully-crafted implementations that
rely on non-trivial technical devices to keep track of causal
dependencies and avoid or resolve conflicts. For instance, various
CRDT variants of the set abstract data type rely on \emph{vector
clocks} to identify conflicting \C{add}s and
\C{remove}s~\cite{zawirski-thesis, zhang}, Replicated Growable Array
(RGA) -- a CRDT for collaborative editing~\cite{rga}, and JSON CRDT --
a CRDT variant of the JSON storage format, both use \emph{Lamport
timestamps}~\cite{rga, json-crdt}; and TreeDoc, another CRDT for
collaborative editing, makes use of \emph{dense linear
orders}~\cite{treedoc}. Such advanced implementations makes it hard to
reason about basic correctness properties of CRDTs, such as
convergence. Indeed, formal verification of \emph{strong eventual
consistency} (i.e., eventual convergence) for few of the
aforementioned CRDTs required considerable effort on behalf of
multiple authors who are experts in the field~\cite{kleppmann2017}.
The extraordinary effort and the expertise required to build CRDTs is
a deterrent towards using type-safe abstractions with strong
guarantees in distributed applications. To overcome this deterrent,
there is a need for an alternative approach that lets developers reuse
their sequential abstractions in a distributed setting with little to
no additional overhead of reasoning about their correctness properties
such as convergence.

% In this paper we propose such an alternative approach to convergent
% replicated data types. The key distinguishing characteristic of our
% approach is the utilization of a distributed runtime to orchestrate
% \emph{well-formed} distributed executions that are guaranteed to
% converge. We define well-formedness as a constraint over the
% \emph{states} that arise during the execution as opposed to the
% \emph{operations} that are involved in the execution. To allow for its
% enforcement, we adopt a state-centric model of replication based on
% version-controlled \emph{mergeable} states~\cite{mrdt} instead of an
% operation-centric model based on commutative operations. Unlike
% commutativity, mergeability does not require a data type definition to
% be refactored to suit distributed execution. 
In this paper we describe an alternative approach to building
convergent replicated data types that realizes the aforementioned
virtues. Our approach is based on Mergeable Replicated Data Types
(MRDTs)~\cite{mrdt} -- an alternative take on RDTs that is inspired by
the Git version control system. MRDTs adopt a state-centric model of
replication based on version-controlled \emph{mergeable} states
instead of an operation-centric model based on commutative operations.
Unlike commutativity, mergeability does not require a data type
definition to be refactored to suit distributed execution. Instead,
the type definition only needs to be extended with a \C{merge}
function to reconcile concurrent versions of the type in presence of
their common ancestor version. However, mergeability itself doesn't
guarantee convergence; there exist MRDTs with well-defined and
intuitive merge semantics that nonetheless admit divergent executions.
While constraining the data type semantics to conform to algebraic
laws such as monotonicity might restore convergence, it would lead
MRDTs down the path of CRDTs, which has known problems described
above. We therefore propose an alternative approach to convergence --
one that is based on runtime orchestration rather than static
enforcement. In particular, we extend MRDTs with a distributed runtime
that orchestrates only the \emph{well-formed} executions which are
guaranteed to converge. Notably, our approach guarantees convergence
of MRDTs \emph{regardless} of their merge semantics, thus obviating
the need to constrain the type semantics or their implementations. The
key insight behind our approach is the observation that a data type's
\C{merge} is a complete function, and thefore can be safely invoked to
merge concurrent versions of the type notwithstanding their individual
operation histories. Thus, a replica's current version
can always be made available for user operations with the guarantee
that it can be safely merged with a concurrent version at a later
point of time. The ability to arbitrarily delay merges gives us the
opportunity to orchestrate them in a way that is \emph{effectively}
linear, which inturn guarantees convergence.
% combined with the simplicity
% of state-based reasoning. Note that state-centric model of
% replication has also been explored in the context of CRDTs. However,
% state-based CRDTs require the replicated state to be organized as a
% lattice -- a restriction we relax in our setting.

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions. Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to \emph{not} interfere with the execution
of RDT operations. Orchestrating a well-formed convergent execution
without impacting latency and availability may seem impossible due to
the CAP theorem~\cite{cap}, but it is not the case in fact. While CAP
theorem does rule out linearizability, it does not prohibit enforcing
weaker consistency constraints on distributed executions. Our approach
exploits this observation by (\rom{1}).  Identifying a novel set of
constraints on distributed executions that guarantee their
convergence, and (\rom{2}). Localizing such constraints on
(asynchronously executed) state merges such that per-operation latency
and system-wide availability remain unaffected. The downside of
interfering with merges, however, is that the replicas may now take
longer to converge -- a phenomenon we quantify using a metric called
\emph{staleness} (Sec.~\ref{sec:eval}). Our approach thus brings to
fore a fundamental tradeoff that was never explored in the context of
RDTs before.

%Sec.~\ref{sec:motivation} explains these insights in detail.

@@

  \item We formalize the aforementioned constraints in the context of
    an abstract machine (named \quark) that manifests distributed
    executions as version history graphs similar to Git.  We prove the
    \emph{convergence} and \emph{progress} properties of the \quark
    abstract machine that together guarantee its soundness. The
    formalization has also been mechanized in Ivy~\cite{ivy} and
    automatically verified with help of Z3~\cite{z3}. To the best of
    our knowledge, this is the first time the meta-theory of a formal
    is proven completely inside an SMT solver.

@@

>>>>

Large-scale web services and decentralized applications often rely on
geo-distributed state replication to meet their latency and availability
needs. An application's state is replicated asynchronously in such a
setting, meaning that operations are independently executed at each
replica, and updates are applied asynchronously at other replicas after a
possible network delay. Asynchronous execution complicates programming and
reasoning about distributed applications as it induces the possibility of
conflicting updates leading to non-convergence and application integrity
violations. Two basic approaches have been proposed to address this
problem. The first approach is to selectively strengthen the system
consistency to pre-empt the conflicting updates. The second is to redefine
the application state in terms of \emph{Replicated Data Types} (RDTs) that
are specially engineered to handle conflicting updates. Strengthening
system consistency necessarily entails inter-replica coordination,
therefore applications prefer to utilize RDTs as much as possible,
resorting to consistency strengthening only when it is necessary to maintain
application integrity. The common design principle guiding the development
of replicated data types is commutativity. The idea is that if the
replicated state is only updated by commutative operations, then updates
can be applied in any order and the replica states are still guaranteed to
converge. Indeed, there exist common use cases in web applications that can
be implemented using Commutative Replicated Data Types
(CRDTs)~\cite{crdts}. For instance, a video view counter on a streaming
application such as YouTube can be implemented as a Counter CRDT that
supports commutative increments.

In general, however, commutativity is not a common occurrence among
data types. Most common data type definitions come with at least a
pair of operations that do not commute. For instance \C{add} and
\C{remove} operations on a set do not commute if both are for the same
element.  Likewise two \C{insert} operations for the same position in
a list do not commute. Such non-commutative operations translate to
conflicting updates in a distributed setting, whose cumulative result
cannot be determined by the sequential specification of the data type
alone. To use a non-commutative data type in a replicated setting,
creative re-engineering of its internal representation and algorithms
is needed to turn it into a bonafide CRDT with a sensible distributed
semantics.
% In the case of set data type, that could involve changing the internal
% representation to maintain two sets -- a set containing the current
% elements and a \emph{tombstone} set that contains every element that
% has ever been deleted from the set. A \C{remove $k$} operation removes
% $k$ from the element set and adds it to the tombstone set, and an
% \C{insert $k$} operation adds $k$ to the element set only if it is not
% present in the tombstone set (i.e., only if $k$ was not previously
% deleted from the set). This results in a CRDT set called
% \emph{two-phase set} where an element once removed cannot be re-added. 
% A more complex re-engineering of the set data type could result in a set
% CRDT with a different (and arguably more useful) semantics.
% Sec.~\ref{sec:motivation} describes a \emph{remove-wins} CRDT set, which
% allows elements to be readded, but which prioritizes \C{remove}s over
% concurrent \C{insert}s. This transformation requires vector
% clocks~\cite{vectorclock} to keep track of casual ordering among
% \C{insert}s and \C{remove}s.
Indeed, most CRDTs have such carefully-crafted implementations that
rely on non-trivial technical devices to keep track of causal
dependencies and avoid or resolve conflicts. For instance, various
CRDT variants of the set abstract data type rely on \emph{vector
clocks} to identify conflicting \C{add}s and
\C{remove}s~\cite{zawirski-thesis, zhang}, Replicated Growable Array
(RGA) -- a CRDT for collaborative editing~\cite{rga}, and JSON CRDT --
a CRDT variant of the JSON storage format, both use \emph{Lamport
timestamps}~\cite{rga, json-crdt}; and TreeDoc, another CRDT for
collaborative editing, makes use of \emph{dense linear
orders}~\cite{treedoc}. Such advanced implementations makes it hard to
reason about basic correctness properties of CRDTs, such as
convergence. Indeed, formal verification of \emph{strong eventual
consistency} (i.e., eventual convergence) for few of the
aforementioned CRDTs required considerable effort on behalf of
multiple authors who are experts in the field~\cite{kleppmann2017}.
The extraordinary effort and the expertise required to build CRDTs is
a deterrent towards using type-safe abstractions with strong
guarantees in distributed applications. To overcome this deterrent,
there is a need for an alternative approach that lets developers reuse
their sequential abstractions in a distributed setting with little to
no additional overhead of reasoning about their correctness properties
such as convergence.

In this paper we describe an alternative approach to building
convergent replicated data types that realizes the aforementioned
virtues. Our approach is based on Mergeable Replicated Data Types
(MRDTs)~\cite{mrdt} -- an alternative take on RDTs that is inspired by
the Git version control system. MRDTs adopt a state-centric model of
replication based on version-controlled \emph{mergeable} states
instead of an operation-centric model based on commutative operations.
Unlike commutativity, mergeability does not require a data type
definition to be refactored to suit distributed execution. Instead,
the type definition only needs to be extended with a \C{merge}
function to reconcile concurrent versions of the type in presence of
their common ancestor version. However, mergeability itself doesn't
guarantee convergence; there exist MRDTs with well-defined and
intuitive merge semantics that nonetheless admit divergent executions.
While constraining the data type semantics to conform to algebraic
laws such as monotonicity might restore convergence, it would burden
the developer with the task of enforcing such constraints, which we
would like to avoid. We therefore propose an alternative approach to
convergence -- one that is based on runtime orchestration rather than
static enforcement. In particular, we extend MRDTs with a distributed
runtime that orchestrates only the \emph{well-formed} executions which
are guaranteed to converge. Notably, our approach guarantees
convergence of MRDTs \emph{regardless} of their merge semantics, thus
obviating the need to constrain the type semantics or their
implementations. The key insight behind our approach is the
observation that a data type's \C{merge} is a complete function, and
thefore can be safely invoked to merge concurrent versions of the type
notwithstanding their individual operation histories. Thus, a
replica's current version can always be made available for user
operations with the guarantee that it can be safely merged with a
concurrent version at a later point of time. The ability to
arbitrarily delay merges gives us the opportunity to orchestrate them
in a way that is \emph{effectively} linear, which inturn guarantees
convergence.

Note that it is easy to construct a trivial runtime that guarantees
convergence of all executions. Such a runtime would make an extensive
use of inter-replica coordination to synchronize the execution of
every operation and thus induce a sequentially-consistent behavior.
This would however defeat the purpose of state replication as the
latency incurred for synchronized execution of an operation is
prohibitively high, and the availability of the system during the
execution is correspondingly low. It is therefore important for a
distributed runtime of RDTs to \emph{not} interfere with the execution
of RDT operations. Orchestrating a well-formed convergent execution
without impacting latency and availability is possible despite the the
CAP theorem~\cite{cap}. While CAP theorem does rule out
linearizability, it does not prohibit enforcing weaker consistency
constraints on distributed executions. Our approach exploits this
observation by (\rom{1}).  Identifying a novel set of constraints on
distributed executions that guarantee their convergence, and
(\rom{2}). Localizing such constraints on (asynchronously executed)
state merges such that per-operation latency and system-wide
availability remain unaffected. The downside of interfering with
merges, however, is that the replicas may now take longer to converge
-- a phenomenon we quantify using a metric called \emph{staleness}
(Sec.~\ref{sec:eval}). Our approach thus brings to fore a fundamental
tradeoff that has not been explored in the context of RDTs.

@@

  \item We formalize the aforementioned constraints in the context of
    an abstract machine (named \quark) that manifests distributed
    executions as version history graphs similar to Git.  We prove the
    \emph{convergence} and \emph{progress} properties of the \quark
    abstract machine that together guarantee its soundness. The
    formalization has also been mechanized in Ivy~\cite{ivy} and
    automatically verified with the help of Z3~\cite{z3}. To the best
    of our knowledge, this is the first time the meta-theory of a
    formal system is proven completely inside an SMT solver.

@@

While our focus in this paper is on RDTs, we believe our contributions
are also meaningful in the context of the long line of work on
programming models for shared-memory concurrency. In particular, this
work generalizes concurrent revisions~\cite{BBL+10} beyond fork-join
parallelism and server-client paradigm. Our approach demonstrates how
causal coherency-based semantics can be implemented more generally in
decentralized systems while guaranteeing the convergence of concurrent
executions.

### 2022-03-21 motivation.tex
Edits for shepherd review and camera-ready

\noindent\paragraph{From Set ADT to Set RDT} Fig.~\ref{fig:ocaml-set}
shows a simplified interface of \C{Set} (abstract) data type in OCaml.
The interface hides a reference to a set (\C{Set.t}), which can be
updated in-place via \C{add} and \C{remove} operations. \C{Set} is an
ordinary data type meant for for sequential execution. Under a
concurrent execution with an asynchronously replicated state, \C{Set}
would exhibit anamolous behaviors such as those in
Fig.~\ref{fig:crdt-execs}.

@@

Extending the \C{Set} implementation from Fig.~\ref{fig:cc-set} with
\emph{add-wins} semantics is however not trivial. The suggested
approach involves tracking element-wise causal dependencies between
the \C{remove}s and \C{add}s by maintaining a vector clock \emph{for
each element} $e$ in the set~\cite{zawirski-thesis}. The vector clock
of $e$ records the sequence number of the last \C{add($e$)} operation
from each replica whose effect has been received and applied at the
current replica. The \C{Add} and \C{Remove} effects on $e$ will now be
tagged with $e$'s vector clock. When an \C{Remove} effect on $e$ is
received at a replica, it is applied only if the tagged vector clock
is no less than $e$'s local vector clock, i.e., only if the arriving
\C{Remove} has seen (i.e., causally succeeds) at least those
\C{add($e$)} operations the current replica is aware of.  Otherwise
the effect is simply a no-op. This strategy would result in convergent
states despite the execution in Fig.~\ref{fig:crdt-exec-2} as $R_2$'s
\C{remove} effectively becomes a no-op at $R_1$ due to there being at
least one \C{add} operation on $e$ that it hasn't seen. The
\emph{add-wins} \C{Set} interface is similar to Fig.~\ref{fig:cc-set},
except that it requires an additional data structure to track
element-wise vector clocks: 
\begin{ocaml}
  val e_clock: (elt, v_clock) HashTable.t
\end{ocaml}
The resultant set RDT is assumed to be correct albeit a formal proof
of convergence could not be found in the literature.

@@

\noindent\paragraph{Well-formed executions and the \quark runtime}
Convergence however is not an inherent virtue of the state-centric
replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
$R_2$ performs a \C{remove} and and an \C{add} making versions
$v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
\C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
merging versions $v_1$ and $v_2$ to make new versions $v_4$ and $v_5$
respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
$R_2$ obtains $R_1$'s changes by merging $v_3$ with $v_2$ (LCA =
$v_0$) to make $v_6$.  Now $R_1$ and $R_2$ have same set of changes
yet their final versions differ.

@@

In the context of Fig.~\ref{fig:mrdt-exec-3}, it is quite clear what
linearization of merges means and how to enforce it via
synchronization (for e.g., wrapping each merge within a global lock).
In general however, the semantics of merge linearization isn't as cut
and dried. For instance, consider the execution in
Fig.~\ref{fig:mrdt-exec-4}. The four replicas involved in the
execution start with version $v_0 \,=\, \{e\}$. The replicas perform
local operations as shown in the figure to make versions $v_1$ to
$v_3$. Next they perform a series of merges to propagate local
changes. The merges can be ordered in time as following: first $R_2
\rightarrow R_1$ (merging $v_1$), then $R_3 \rightarrow R_1$ twice
(merging $v_2$ and $v_3$ resp.), then $R_3 \rightarrow R_4$ (merging
$v_3)$, and finally $R_2 \rightarrow R_4$ (merging $v_1$). These
merges collectively propagate the effects of \C{add} and \C{remove}
operations to $R_1$ and $R_2$. And despite being ordered in time,
they nonetheless result in a divergent execution ($v_5 \neq v_7$). The
problem here is that, although merges are executed linearly, the
execution graph does not reflect this linearity; merges that end in
$R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
those that end in $R_4$. This shows that simply synchronizing the
execution of merges does \emph{not} result in convergence.

Our key insight to overcome this impasse is a novel \emph{well-formedness}
condition on execution graphs that ensures convergence of final
states. To understand well-formedness, let us contrast the bad
executions in Figs.~\ref{fig:mrdt-exec-2} and~\ref{fig:mrdt-exec-4}
against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
in Fig.~\ref{fig:mrdt-exec-3}, \emph{every} pair of concurrent
versions on $R_1$ and $R_2$ have a unique lowest common ancestor
(LCA). For instance, LCA of $(v_5,v_6)$ is $v_5$, $(v_5,v_4)$ is
$v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
Fig.~\ref{fig:mrdt-exec-2}, versions $v_5$ and $v_6$ have two LCAs,
namely $v_2$ and $v_3$. Both these versions are common ancestors of
$v_5$ and $v_6$, and both are \emph{lowest} in the sense that there do
not exist versions lower (in the execution graph) than $v_2$ and $v_3$
that are also common ancestors of $v_5$ and $v_6$. Likewise in
Fig.~\ref{fig:mrdt-exec-4}, versions $v_6$ and $v_8$ have two LCAs --
$v_1$ and $v_3$. Multiple LCAs is an indication that there exist
merges prior in the execution graph that are effectively concurrent.
Considering that our approach to convergence in state-centric model
crucially relies on linearity of merges, presence of multiple LCAs
opens up the possibility of divergence. We therefore define
well-formed execution graphs as those where LCAs for every pair of
concurrent versions is unique. As we prove in
Sec.~\ref{sec:abstract-sem}, enforcing this structural
well-formedness condition indeed guarantees the convergence of
distributed executions. In Sec.~\ref{sec:concrete-sem}, we formalize a
distributed runtime called \quark whose primary purpose is to enforce
the well-formedness condition on distributed executions. In
Sec.~\ref{sec:implementation}, we describe its implementation. The
runtime makes it possible to automatically enforce convergence,
letting us promote ordinary sequential data types to convergent
replicated types by simply equipping them with a merge function.
Thus, extending the \C{Set} interface of Fig.~\ref{fig:ocaml-set} with
the two-line \C{Set.merge} function shown above is sufficient to
obtain a convergent replicated set data type.

>>>>

\noindent\paragraph{From Set ADT to Set RDT} Fig.~\ref{fig:ocaml-set}
shows a simplified interface of \C{Set} (abstract) data type in OCaml.
The interface hides a reference to a set (\C{Set.t}), which can be
updated in-place via \C{add} and \C{remove} operations. \C{Set} is an
ordinary data type meant for sequential execution. Under a
concurrent execution with an asynchronously replicated state, \C{Set}
would exhibit anamolous behaviors such as those in
Fig.~\ref{fig:crdt-execs}.

@@

Extending the \C{Set} implementation from Fig.~\ref{fig:cc-set} with
\emph{add-wins} semantics is however not trivial. The suggested
approach involves tracking element-wise causal dependencies between
the \C{remove}s and \C{add}s by maintaining a vector clock \emph{for
each element} $e$ in the set~\cite{zawirski-thesis}. The vector clock
of $e$ records the sequence number of the last \C{add($e$)} operation
from each replica whose effect has been received and applied at the
current replica. The \C{Add} and \C{Remove} effects on $e$ will now be
tagged with $e$'s vector clock. When a \C{Remove} effect on $e$ is
received at a replica, it is applied only if the tagged vector clock
is no less than $e$'s local vector clock, i.e., only if the arriving
\C{Remove} has seen (i.e., causally succeeds) at least those
\C{add($e$)} operations the current replica is aware of.  Otherwise
the effect is simply a no-op. This strategy would result in convergent
states despite the execution in Fig.~\ref{fig:crdt-exec-2} as $R_2$'s
\C{remove} effectively becomes a no-op at $R_1$ due to there being at
least one \C{add} operation on $e$ that it hasn't seen. The
\emph{add-wins} \C{Set} interface is similar to Fig.~\ref{fig:cc-set},
except that it requires an additional data structure to track
element-wise vector clocks: 
\begin{ocaml}
  val e_clock: (elt, v_clock) HashTable.t
\end{ocaml}
The resultant set RDT is assumed to be correct albeit a formal proof
of convergence could not be found in the literature.

@@

\noindent\paragraph{Well-formed executions and the \quark runtime}
Convergence however is not an inherent virtue of the state-centric
replication model. Fig.~\ref{fig:mrdt-exec-2} shows the state-centric
analogue of the execution in Fig.~\ref{fig:crdt-exec-2} which
diverges. Here $R_1$ and $R_2$ start with version $v_0 \,=\, \{e\}$.
$R_2$ performs a \C{remove} and an \C{add} making versions
$v_1$ and $v_2$ respectively. Simultaneously, $R_1$ performs a
\C{remove} to make $v_3$. Replica $R_1$ now obtains $R_2$'s changes by
merging versions $v_1$ and $v_2$ to make new versions $v_4$ and $v_5$
respectively. LCAs for these merges are $v_0$ and $v_1$. Concurrently,
$R_2$ obtains $R_1$'s changes by merging $v_3$ with $v_2$ (LCA =
$v_0$) to make $v_6$.  Now $R_1$ and $R_2$ have same set of changes
yet their final versions differ.

@@

In the context of Fig.~\ref{fig:mrdt-exec-3}, it is quite clear what
linearization of merges means and how to enforce it via
synchronization (for e.g., wrapping each merge within a global lock).
In general however, the semantics of merge linearization isn't as cut
and dried. For instance, consider the execution in
Fig.~\ref{fig:mrdt-exec-4}. The four replicas involved in the
execution start with version $v_0 \,=\, \{e\}$. The replicas perform
local operations as shown in the figure to make versions $v_1$ to
$v_3$. Next they perform a series of merges to propagate local
changes. The merges can be ordered in time as following: first $R_2
\rightarrow R_1$ (merging $v_1$), then $R_3 \rightarrow R_1$ twice
(merging $v_2$ and $v_3$ resp.), then $R_3 \rightarrow R_4$ (merging
$v_3)$, and finally $R_2 \rightarrow R_4$ (merging $v_1$). These
merges collectively propagate the effects of \C{add} and \C{remove}
operations to $R_1$ and $R_4$. And despite being ordered in time,
they nonetheless result in a divergent execution ($v_6 \neq v_8$). The
problem here is that, although merges are executed linearly, the
execution graph does not reflect this linearity; merges that end in
$R_1$ in Fig.~\ref{fig:mrdt-exec-4} are effectively concurrent with
those that end in $R_4$. This shows that simply synchronizing the
execution of merges does \emph{not} result in convergence.

Our key insight to overcome this impasse is a novel \emph{well-formedness}
condition on execution graphs that ensures convergence of final
states. To understand well-formedness, let us contrast the bad
executions in Figs.~\ref{fig:mrdt-exec-2} and~\ref{fig:mrdt-exec-4}
against the good execution in Fig.~\ref{fig:mrdt-exec-3}. Observe that
in Fig.~\ref{fig:mrdt-exec-3}, \emph{every} pair of concurrent
versions on $R_1$ and $R_2$ have a unique lowest common ancestor
(LCA). For instance, LCA of $(v_5,v_6)$ is $v_5$, $(v_5,v_4)$ is
$v_3$, $(v_1,v_2)$ is $v_0$ and so on. By contrast in
Fig.~\ref{fig:mrdt-exec-2}, versions $v_5$ and $v_6$ have two LCAs,
namely $v_2$ and $v_3$. Both these versions are common ancestors of
$v_5$ and $v_6$, and both are \emph{lowest} in the sense that there do
not exist versions lower (in the execution graph) than $v_2$ and $v_3$
that are also common ancestors of $v_5$ and $v_6$. Likewise in
Fig.~\ref{fig:mrdt-exec-4}, versions $v_6$ and $v_8$ have two LCAs --
$v_1$ and $v_3$. Multiple LCAs is an indication that there exist
merges prior in the execution graph that are effectively concurrent.
Considering that our approach to convergence in the state-centric
model crucially relies on the linearity of merges, presence of
multiple LCAs opens up the possibility of divergence. We therefore
define well-formed execution graphs as those where LCAs for every pair
of concurrent versions is unique. As we prove in
Sec.~\ref{sec:abstract-sem}, enforcing this structural well-formedness
condition indeed guarantees the convergence of distributed executions.
In Sec.~\ref{sec:concrete-sem}, we formalize a distributed runtime
called \quark whose primary purpose is to enforce the well-formedness
condition on distributed executions. In Sec.~\ref{sec:implementation},
we describe its implementation. The runtime makes it possible to
automatically enforce convergence, letting us promote ordinary
sequential data types to convergent replicated types by simply
equipping them with a merge function.  Thus, extending the \C{Set}
interface of Fig.~\ref{fig:ocaml-set} with the two-line \C{Set.merge}
function shown above is sufficient to obtain a convergent replicated
set data type.

### 2022-03-21 quark.tex
Edits for shepherd review and camera-ready

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
\documentclass[sigplan,screen]{acmart}
%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% use defaults for review submission.
%%\acmJournal{PACMPL}
%%\acmVolume{1}
%%\acmNumber{POPL} % CONF = POPL or ICFP or OOPSLA
%%\acmArticle{1}
%%\acmYear{2022}
%%\acmMonth{1}
%%\acmDOI{} % \acmDOI{10.1145/nnnnnnn.nnnnnnn}
%%\startPage{1}
\acmConference[PLDI]{ACM Conference}{June 2022}{San Diego, USA}%

%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\copyrightyear{2018}           %% If different from \acmYear

%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmnumeric}   %% For author/year citations

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data
  type operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.
  The novel component of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of latency and availability. Instead, we
  introduce a novel tradeoff against a metric called \emph{staleness},
  which roughly corresponds to the time taken for replicas to
  converge. We implement our approach in a system called \quark and
  conduct a thorough evaluation of its tradeoffs in the context of two
  case studies.
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Collaborative
Editing}  %% \keywords are mandatory in final camera-ready submission

@@

%% Acknowledgments
%\begin{acks}                            %% acks environment is optional
                                        %% contents suppressed with 'anonymous'
  %% Commands \grantsponsor{<sponsorID>}{<name>}{<url>} and
  %% \grantnum[<url>]{<sponsorID>}{<number>} should be used to
  %% acknowledge financial support and will be used by metadata
  %% extraction tools.
%\end{acks}

>>>>

%% For double-blind review submission, w/o CCS and ACM Reference (max submission space)
%\documentclass[sigplan,review]{acmart}
\documentclass[sigplan,screen]{acmart}
%\settopmatter{printfolios=true,printccs=false,printacmref=false}
%% For final camera-ready submission, w/ required CCS and ACM Reference
%\documentclass[acmsmall]{acmart}\settopmatter{}

@@

%% Journal information
%% Supplied to authors by publisher for camera-ready submission;
%% Copyright information
%% Supplied to authors (based on authors' rights management selection;
%% see authors.acm.org) by publisher for camera-ready submission;
%% use 'none' for review submission.
%%\setcopyright{none}
\copyrightyear{2022} 
\acmYear{2022} 
\setcopyright{rightsretained} 
\acmConference[PLDI '22]{Proceedings of the 43rd ACM SIGPLAN International Conference on Programming Language Design and Implementation}{June 13--17, 2022}{San Diego, CA, USA}
\acmBooktitle{Proceedings of the 43rd ACM SIGPLAN International Conference on Programming Language Design and Implementation (PLDI '22), June 13--17, 2022, San Diego, CA, USA}
\acmDOI{10.1145/3519939.3523724}
\acmISBN{978-1-4503-9265-5/22/06}
%% Bibliography style
\bibliographystyle{ACM-Reference-Format}
%% Citation style
%% Note: author/year citations are required for papers published as an
%% issue of PACMPL.
\citestyle{acmnumeric}   %% For author/year citations

@@

%% Abstract
%% Note: \begin{abstract}...\end{abstract} environment must come
%% before \maketitle command
\begin{abstract}
  We propose a runtime-assisted approach to enforce convergence in
  distributed executions of replicated data types. The key
  distinguishing aspect of our approach is that it guarantees
  convergence \emph{unconditionally} -- without requiring data type
  operations to satisfy algebraic laws such as commutativity and
  idempotence. Consequently, programmers are no longer obligated to
  prove convergence on a per-type basis. Moreover, our approach lets
  sequential data types be reused in a distributed setting by
  extending their implementations rather than refactoring them.  The
  novel component of our approach is a distributed runtime that
  orchestrates \emph{well-formed } executions that are guaranteed to
  converge. Despite the utilization of a runtime, our approach comes
  at no additional cost of latency and availability. Instead, we
  introduce a novel tradeoff against a metric called \emph{staleness},
  which roughly corresponds to the time taken for replicas to
  converge. We implement our approach in a system called \quark and
  conduct a thorough evaluation of its tradeoffs in the context of two
  case studies. 
%%We also briefly discuss the utility of our approach in
%%realizing causal coherency semantics in shared-memory concurrent
%%systems,
\end{abstract}

@@

%% Keywords
%% comma separated list
\keywords{Replication, MRDT, CRDT, Runtime, Convergence, Concurrent
Revisions, Causal Consistency, decentralized systems}  %% \keywords are mandatory in final camera-ready submission

@@

%% Acknowledgments
\begin{acks}                            %% acks environment is optional
  We thank the anonymous reviewers and our shepherd Doug Lea for their
  thoughtful comments. First author has benefited from substantial
  discussions with Anmol Sahoo and Suresh Jagannathan. Anmol has also
  developed and maintained the OCaml-Scylla library, which is a key
  piece of infrastructure underlying \quark.
\end{acks}
